---
title: "Lab 11"
author: "McKenzie Mai (A16646664)"
date: "2024-05-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1. Proportion og G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39580100-40210101;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```


```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) * 100
```

Now let's look at a different population. I  picked the GBR

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```


Find proportion of G|G
```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) *100, 2)
```


This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Let's now dig into this further.

## Section 4: Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
# sample size:
table(expr$geno)
```

```{r}
# median expression levels:
library(dplyr)
expr %>%
  group_by(geno)%>%
  summarize(median_expression=median(exp))
```

```{r}
library(ggplot2)
```


Lets make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

boxplot shown above; A/A is more expressed in comparison to G/G because it has a higher median expression value. yes, the SNP affects the expression because expression is seen to decreases as each allele changes.


