---
title: "Class 5: Data Visualization with ggplot"
author: "McKenzie Mai (A16646664)"
format: pdf
toc: true
---

Today we will have our first play with the **ggplot2** package - one of the most popular graphics packages on the planet.

There are many plotting systems in R. These include so-called *"base"* plotting/graphics

```{r}
plot(cars)
```

Base plot is generally rather short code and somewhat dull plots - but it is always there for you and is fast for big datasets.


If I want to use **ggplot2** it takes some more work.

```{r}
# ggplot(cars)
```

I need to install the package first to my computer. To do this I can use the function `install.packages("ggplot2")`

Every time I want to use a package I need to load it up with a `library()` call.


```{r}
library(ggplot2)
```


Now finally I can use ggplot

```{r}
ggplot(cars)
```

Every ggplot has at least 3 things:

- **data** (the data.frame with the data you want to plot)
- **aes** (the aesthetic mapping of the data to the plot)
- **geom** (how do you want the plot to look, points, lines, etc.)

```{r}
head(cars)
```

```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth()
```


I want a linear model and no standard error bounds shown on my plot. I also want nicer axis labels and title etc.

```{r}
bp <- ggplot(cars) + aes(x=speed, y=dist) + geom_point()
```

```{r}
bp
```

```{r}
bp + geom_smooth()
```

```{r}
bp + geom_smooth(se=FALSE, method="lm") +
  labs(title="Stopping Distance of Old Cars",
       x= "Speed (MPH)",
       y= "Distance(ft)",
       caption= "From the cars dataset") +
  theme_bw()
```

## A more complicated scatter plot

Here we make a plot of gene expression data:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)

```

```{r}
head(genes)
```

Q. Use the nrow() function to find out how many genes are in this dataset. What is your answer?
```{r}
nrow(genes)
```

```{r}
colnames(genes)
ncol(genes)
```

Q. Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer?
```{r}
table(genes$State)
```

```{r}
round( sum(genes$State == "up") /nrow(genes) * 100, 2)

```

```{r}
n.gene <- nrow(genes)
n.up <- sum(genes$State == "up")

up.percent <-  n.up/n.gene * 100
round(up.percent, 2)
```

```{r}
head(genes, 2)
```

```{r}
p <- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
p
```

```{r}
p + labs(title="Gene Expression Changes on Drug Treatment",
         x="Control (no drug)",
         y="Treated (drug)") +
  scale_colour_manual( values=c("blue","gray","red") )
```

## Exploring the gapminder dataset

Here we will load up the gapminder dataset to get practice with different aes mappings.

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

> Q. How many entries rows are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many columns?

```{r}
ncol(gapminder)
```

```{r}
head(gapminder)
```

```{r}
table(gapminder$year)
```

> Q. How many continents?

```{r}
table(gapminder$continent)
```

I could use the `unique()` function...

```{r}
length(unique(gapminder$continent))
```

> Q. How many countries are there in this dataset?

```{r}
length(unique(gapminder$country))
```

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp)+
  geom_point(alpha=0.2, col="blue")
```

```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

head(gapminder_2007)
```

Plot of 2007 with population and continent data
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, col= continent, size=pop) +
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp)+
  geom_point() +
  facet_wrap(~continent)
```

