---
title: "Class 7: Machine learning 1"
author: "McKenzie Mai (A16646664)"
format: gfm
---

Today we will start our multi-part exploration of some key machine learning methods. We will begin with clustering- finding groupings in data, and then dimensionallity reduction.


## Clustering

Let's start with "k-means" clustering.
The main function in base R for this `kmeans()`

```{r}
# Make up some data
hist( rnorm(100000, mean=3) )
```


```{r}
tmp <- c(rnorm(30, -3), rnorm(30, +3))
x <- cbind(x= tmp, y=rev(tmp))
plot(x)
```

Now let's try out `kmeans()`

```{r}
km <- kmeans(x, centers=2)
km
```

```{r}
attributes(km)
```

> Q. How many popints in each cluster?

```{r}
km$size
```

> Q. What component of your result object details cluster assignment/membership?

```{r}
km$cluster
```

> Q. What are centers/mean values of each cluster?

```{r}
km$centers
```


> Q. Make a plot of your data showing your clustering results (groupings/clusters and cluster centers).

```{r}
plot(x, col=c("red", "blue"))
```


```{r}
c(1:5) + c(100,1)
```

```{r}
plot(x, col=c(1,2))
```


```{r}
plot(x, col=km$cluster)
points(km$centers, col="green", pch=15, cex=3)
```


> Q. Run `kmeans()` agan and cluster in 4 groups and plot the results.

```{r}
km4 <- kmeans(x, centers = 4)
plot(x, col=km4$cluster)
```

## Hierarchical Clustering

This form of clustering aims to reveal the structure in your data by progressively grouping points into an even smaller number of clusters.

The main function in base R for this called `hclust()`. This function does not take our input data directly but wants a "distance matrix" that details how (dis)similar all our input points are to each other.


```{r}
hc <- hclust(dist(x))
hc
```

The print out above is not very useful (unlike that from kmeans) but there is a useful `plot()` method.

```{r}
plot(hc)
abline(h=10, col="red")
```

To get my main result (my cluster membership vector) I need to "cut" my tree using the function `cutree()`

```{r}
grps <- cutree(hc, h=10)
grps
```

```{r}
plot(x, col=grps)
```



```{r}
plot(x, col=cutree(hc, h=6))
```

# Principal Component Analysis (PCA)

The goal of PCA is to reduce the dimensionality of a dataset down to some smaller subset of new variables (called PCs) that are a useful bases for further analysis, like visualization, clustering, etc.

## Data import

Read data about crazy eating trends in the UK and N. Ireland

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```



```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

The so-called "pairs" plot can be useful for small datasets:

```{r}
#rainbow(nrow(x))
pairs(x, col=rainbow(nrow(x)), pch=16)
```

So the paris plot is useful for small datasets but it can be lots of work to interpret and gets intractable for larger datasets.

so PCA to the rescue...


The main function to do PCA in base R is called `prcomp()`. This function wants the transpose of our data in this case.

```{r}
pca <- prcomp(t(x))
summary(pca)
```

```{r}
attributes(pca)
```

```{r}
pca$x
```

A major PCA result viz is called a "PCA plot" (a.k.a a score plot, a bipolot, PC1 vs PC2 plot, ordination plot)

```{r}
mycols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=mycols, pch=16, 
     xlab="PC1", ylab="PC2")
abline(h=0, col="gray")
abline(v=0, col="gray")
```
Another important output from PCA is called the "loadings" vector or the "rotation" component- this tell sus how much the original variables ( the foods in this case) contributre to the new PCs.

```{r}
pca$rotation
```

PCA looks to be a super useful method for gaining some insight into high dimensional data that is difficult to examine in other ways.


# PCA of RNASeq data

## Data input

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)
```

```{r}
summary(pca)
```

> Q. How many genes in this dataset

```{r}
nrow(rna.data)
```





```{r}
head(pca$x)
```

I will make a main result figure use ggplot:

```{r}
library(ggplot2)
```


```{r}
res <- as.data.frame(pca$x)
```


```{r}
ggplot(res)+
  aes(PC1, PC2) +
  geom_point()
```

```{r}
mycols <- c(rep("blue", 5), rep("red", 5))
mycols
```

```{r}
ggplot(res) +
  aes(x=PC1, y=PC2, label=row.names(res)) +
  geom_point(col=mycols) +
  geom_label(col=mycols)
```

```{r}
colnames(rna.data)
```


```{r}
kmeans(pca$x[,1], centers=2)
```

